Distance Sampling in \proglang{R}

### 2015-12-12 paper.Rmd
Minor changes to manuscript
Small wording changes:  this line doesn't sound right to me, but didn't
know how to correct:

"where $k$ is the key function and $\alpha_O$ is sum series of functions

* In the simplest case, where the objective is to estimate a detection function, all that is needed is a vector of distances.
* To include additional covariates into the detection function (see "Detection functions") we use a \code{data.frame}. The \code{data.frame} contains a column called \code{distance} (containing the observed distances) and additional named columns for covariates potentially useful in modelling detectability (for example \code{observer} or \code{seastate}). Some column names are reserved: \code{object} for an observation identifier (see "Extensions"), \code{size} for group or cluster size (see "Detection functions" and "Abundance and variance estimation"), \code{detected} for whether an observation was detected (see "Extensions") and the columns described in the next bullet.
* To estimate abundance beyond area covered by expended survey effort, additional information is required. This consists of information about which transect and occasion the observation was made (the \code{Sample.Label}), a column named \code{Effort} which gives the effort associated with that sample (for lines their length and for points the number of times that point was visited), the stratum the sample was located in (the \code{Region.Label}, which may be from pre- or post-survey stratification, see "Estimating abundance and variance") and the stratum areas (\code{Area}). For transects which were surveyed but have no observations we enter \code{NA} in their \code{distance} column. We refer to this data format (where all information is contained in one table) as "flatfile" as it can be easily created in a single spreadsheet.

@@

We will explore the effects of including effort and stratum data during analysis below.

@@

It is crucial that the detection function accurately models detectability at small distances; we are less worried by its behaviour further away from 0. To ensure that the model is not overly influenced by distances far from zero, we truncate the distances beyond a given distance $w$ (known as the *truncation distance*). Including these larger distances in out analysis does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, sections 4.3 and 5.3].

@@

* *Shoulder*: we expect observers to be able to see objects near them, not just those directly in front of them. For this reason, we expect the detection function to be flat near the line or point.
* *Non-increasing*: we do not think that observers should be more likely to see distant objects than those nearer the transect. If this occurs, it usually indicates an issue with field procedure (that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
* *Model robust*: models should be flexible enough to make many different shapes.
* *Pooling robust*: many factors can affect the probability of detection and it is not possible to measure all of these. We would like models to produce unbiased results without inclusion of these factors.
* *Estimator efficiency*: we would like models to have low variances, but only given they satisfy the other properties above (which, if they are satisfied, would give low bias).

@@

There is a wide literature on possible formulations for the detection function [@Buckland:1992fa; @Eidous:2005bj; @Becker:2009cj; @Giammarino:2014eg; @Miller:2015hw; @Becker:2015fi]. \code{Distance} includes the most popular of these models. Here we will detail the most popular detection function approach: "key function plus adjustments" (K+A).

@@

Including covariates has an important implication for our calculation of detectability. Because we do not know what the true distribution of the covariates is we must calculate the probability of detection conditional on the observed values of the covariates:
$$
p(\mathbf{z_i}) = \int_0^w \pi(y) g(y, \mathbf{z_i}; \boldsymbol{\theta}) dy,
$$
where $\mathbf{z_i}$ is the vector of $J$ covariates associated with observation $i$. For covariate models, we are calculating a value of "average" probability of detection (average in the sense of distance) per observation. There will be as many unique values of $p(\mathbf{z_i})$ as there are unique covariate combinations in our data.

@@

We can fit a model to the minke whale data, setting the truncation at 1.5km and using the default options in \code{ds} very simply:
```{r minke-hn}
minke_hn <- ds(minke, truncation=1.5)
```
Note that \code{ds} will automatically select adjustment terms by AIC and shows the selection steps.

@@

As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we can use \code{summary} to give useful information about our fitted model. For example for our hazard-rate model for the amakihi, with observer as a covariate:
```{r amakihi-summary}
summary(amakihi_hr_obs)
```
This summary information includes details of the data and model specification, as well as the values of the coefficients ($\beta_j$) and their uncertainties, an "average" value for the detectability (see "Estimating abundance and variance" for details on how this is calculated) and its uncertainty. The final line gives an estimate of abundance for the area covered by the survey (covered in the next section).

@@

To judge goodness of fit for detection functions when exact distances are used, we want to compare the cumulative distribution function (CDF) and empirical distribution function (EDF) for the detection function via a quantile-quantile plot (Q-Q plot). The CDF evaluates the probability of observing an object at a distance less than or equal to some given value. The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance. This can be interpreted as assessing whether the number of observations up to a given distance in line with what the model says they should be (where the "given values" are the observed distances). As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show more deviations from that line.

@@

We can generate a Q-Q plot and test results using the \code{gof_ddf} function. Figure \ref{amakihi-qq} shows the goodness of fit tests for two models for the amakihi data. We first fitting a half-normal model without covariates or adjustments (not setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments):
```{r amakihi-gof, fig.keep="none"}
amakihi_hr <- ds(amakihi, truncation=82.5, transect="point", key="hn", adjustment=NULL)
gof_ds(amakihi_hr)
gof_ds(amakihi_hr_obs_mas)
```
The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}, comparing the above model with the hazard-rate model with observer and minutes after sunrise included.

@@

Once we have a set of models which fit well, we can use Akaike's Information Criterion (AIC) to select between models. \pkg{Distance} includes a function to create table of summary information for fitted models, making it easy to get an overview of a large number of models at once. The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication [@knitr-pkg]. An example of this output is shown in Table \ref{tab:amakihi} and was generated by the following call to \code{summarize_ds_models}:
```{r summary-table-null, eval=FALSE}
summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas)
```

@@

Having obtained the abundance in the covered area, we can then scale-up to the study area:
$$
\hat{N} = \frac{A}{a} \hat{N}_\text{C},
$$
where $A$ is the area of the study region to extrapolate the abundance estimate to and $a$ is the covered area. For line transects $a=2wL$ (twice the truncation distance multiplied by the total length of transects surveyed, $L$) and for points $a=\pi w^2 T$ (where $\pi w^2$ is the area of a single surveyed circle and $T$ is the total number of visits to the sampled points).

We can use the Horvitz-Thompson-like estimator to calculate the "average" detectability for models which include covariates. We can consider what single detectability value would give the estimated $\hat{N}$ and therefore calculate:
$$
\hat{P_a} = n/\hat{N}_\text{C}.
$$
This can be a useful summary statistic, giving us an idea of how detectable our $n$ observed animals would have to be in order to estimate the same $\hat{N}$ if there were no observed covariates. It can also be compared to similar estimates in mark-recapture studies. $P_a$ is included in the \code{summary} output and the table produced by \code{summarize_ds_models}.

@@

We may wish to calculate abundance estimates for some sub-regions of the study region, we call these areas *strata*. For example, strata may be defined by habitat types or animal gender (or some combination) which may be of interest for biological or management reasons. To calculate estimates for a given stratification each observation must occur in a stratum which must be labelled with a \code{Region.Label} and have a corresponding \code{Area} (if we are using an animal characteristic like gender, we would have the areas be the same but if we were using say forested vs. wetland habitat the areas of those strata would be different). Finally, we must also know which stratum each observation occurs in.

>>>>

* In the simplest case, where the objective is to estimate a detection function, all that is needed is a vector of distances.
* To include additional covariates into the detection function (see "Detection functions") we use a \code{data.frame}. The \code{data.frame} contains a column called \code{distance} (containing the observed distances) and additional named columns for covariates potentially useful in modelling detectability (for example \code{observer} or \code{seastate}). Some column names are reserved: \code{object} for an observation identifier (see "Extensions"), \code{size} for group or cluster size (see "Detection functions" and "Abundance and variance estimation"), \code{detected} for whether an observation was detected (see "Extensions") and the columns described in the next bullet.
* To estimate abundance beyond area covered by expended survey effort, additional information is required. This consists of information about transect and occasion the observation was made (the \code{Sample.Label}), a column named \code{Effort} which gives the effort associated with that sample (for lines their length and for points the number of times that point was visited), the stratum the sample was located in (the \code{Region.Label}, which may be from pre- or post-survey stratification, see "Estimating abundance and variance") and the stratum areas (\code{Area}). For transects which were surveyed but have no observations we enter \code{NA} in their \code{distance} column. We refer to this data format (where all information is contained in one table) as "flatfile" as it can be easily created in a single spreadsheet.

@@

We will explore the consequences of including effort and stratum data during analysis below.

@@

It is crucial that the detection function accurately models detectability at small distances; we are less worried by its behaviour further away from 0. To ensure that the model is not overly influenced by distances far from zero, we truncate the distances beyond a given distance $w$ (known as the *truncation distance*). Including these larger distances in our analysis does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, sections 4.3 and 5.3].

@@

* *Shoulder*: we expect observers to be able to see objects near them, not just those directly in front of them. For this reason, we expect the detection function to be flat near the line or point.
* *Non-increasing*: we do not think that observers should be more likely to see distant objects than those nearer the transect. If this occurs, it usually indicates an issue with field procedure (that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
* *Model robust*: models should be flexible enough to make many different shapes.
* *Pooling robust*: many factors can affect the probability of detection and it is not possible to measure all of these. We would like models to produce unbiased results without inclusion of these factors.
* *Estimator efficiency*: we would like models to have low variances, given they satisfy the other properties above (which, if satisfied, would give low bias).

@@

There is a wide literature on possible formulations for the detection function [@Buckland:1992fa; @Eidous:2005bj; @Becker:2009cj; @Giammarino:2014eg; @Miller:2015hw; @Becker:2015fi]. \code{Distance} includes the most popular of these models. Here we detail the most popular detection function approach: "key function plus adjustments" (K+A).

@@

Including covariates has an important implication for our calculation of detectability. Because we do not know  the true distribution of the covariates,we must calculate the probability of detection conditional on the observed values of the covariates:
$$
p(\mathbf{z_i}) = \int_0^w \pi(y) g(y, \mathbf{z_i}; \boldsymbol{\theta}) dy,
$$
where $\mathbf{z_i}$ is the vector of $J$ covariates associated with observation $i$. For covariate models, we are calculating a value of "average" probability of detection (average in the sense of distance) per observation. There will be as many unique values of $p(\mathbf{z_i})$ as there are unique covariate combinations in our data.

@@

We can fit a model to the minke whale data, setting the truncation at 1.5km and using the default options in \code{ds} very simply:
```{r minke-hn}
minke_hn <- ds(minke, truncation=1.5)
```
Note that \code{ds} will automatically select adjustment terms by AIC and show the selection steps.

@@

As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we can use \code{summary} to give useful information about our fitted model. For our hazard-rate model for the amakihi, with observer as a covariate:
```{r amakihi-summary}
summary(amakihi_hr_obs)
```
This summary information includes details of the data and model specification, as well as the values of the coefficients ($\beta_j$) and their uncertainties, an "average" value for the detectability (see "Estimating abundance and variance" for details on how this is calculated) and its uncertainty. The final line gives an estimate of abundance for the area covered by the survey (see the next section).

@@

To judge goodness of fit for detection functions when exact distances are used, we want to compare the cumulative distribution function (CDF) and empirical distribution function (EDF) for the detection function via a quantile-quantile plot (Q-Q plot). The CDF evaluates the probability of observing an object at a distance less than or equal to some value. The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance. This can be interpreted as assessing whether the number of observations up to a given distance in line with what the model says they should be (where the "given values" are the observed distances). As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show greater deviations from that line.

@@

We can generate a Q-Q plot and test results using the \code{gof_ddf} function. Figure \ref{amakihi-qq} shows the goodness of fit tests for two models for the amakihi data. We first fitting a half-normal model without covariates or adjustments (setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments):
```{r amakihi-gof, fig.keep="none"}
amakihi_hr <- ds(amakihi, truncation=82.5, transect="point", key="hn", adjustment=NULL)
gof_ds(amakihi_hr)
gof_ds(amakihi_hr_obs_mas)
```
The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}, comparing the above model with the hazard-rate model with observer and minutes after sunrise included.

@@

Once we have a set of models which fit well, we can use Akaike's Information Criterion (AIC) to select between models. \pkg{Distance} includes a function to create table of summary information for fitted models, making it easy to get an overview of a large number of models. The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication [@knitr-pkg]. An example of this output is shown in Table \ref{tab:amakihi} and was generated by the following call to \code{summarize_ds_models}:
```{r summary-table-null, eval=FALSE}
summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas)
```

@@

Having obtained the abundance in the covered area, we can then scale-up to the study area:
$$
\hat{N} = \frac{A}{a} \hat{N}_\text{C},
$$
where $A$ is the area of the study region to which to extrapolate the abundance estimate and $a$ is the covered area. For line transects $a=2wL$ (twice the truncation distance multiplied by the total length of transects surveyed, $L$) and for points $a=\pi w^2 T$ (where $\pi w^2$ is the area of a single surveyed circle and $T$ is the total number of visits to the sampled points).

We can use the Horvitz-Thompson-like estimator to calculate the "average" detectability for models which include covariates. We can consider what single detectability value would give the estimated $\hat{N}$ and therefore calculate:
$$
\hat{P_a} = n/\hat{N}_\text{C}.
$$
This can be a useful summary statistic, giving us an idea of how detectable our $n$ observed animals would have to be to estimate the same $\hat{N}$ if there were no observed covariates. It can also be compared to similar estimates in mark-recapture studies. $P_a$ is included in the \code{summary} output and the table produced by \code{summarize_ds_models}.

@@

We may wish to calculate abundance estimates for some sub-regions of the study region, we call these areas *strata*. For example, strata may be defined by habitat types or animal gender (or some combination) which may be of interest for biological or management reasons. To calculate estimates for a given stratification each observation must occur in a stratum which must be labelled with a \code{Region.Label} and have a corresponding \code{Area} (if we are using an animal characteristic like gender, we would have the areas be the same but if we were using say forested vs. wetland habitat the areas of those strata would be different). Finally, we must also know the stratum in which each observation occurs.
